---
title: "Week 3: Intro to Likelihood Inference"
subtitle: "PS 818 - Statistical Models"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "September 8, 2025"
date-format: "MMMM D, YYYY"
editor: 
  render-on-save: true
format: 
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 818 - University of Wisconsin - Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute: 
  cache: false
code:
  execute: 
    - r
    - python
embed-resources: true

---

$$
  \require{cancel}
  \DeclareMathOperator*{\argmin}{arg\,min}
$$


$$
 \DeclareMathOperator*{\argmax}{arg\,max}
$$

```{r, echo = F, message=F, warnings=F}
library(tidyverse)
library(haven)
library(estimatr)
library(knitr)
library(broom)
options(digits=3)
```

# Previously

::: {.incremental}
- Linear regression
  - "Agnostic" regression and the "Best Linear Predictor"
  - "Classical" OLS regression theory

- Key intuitions
  - Regression as a **projection** of $Y$ into the column space of $\mathbf{X}$
  - Frisch-Waugh-Lovell theorem

:::

# This week

::: {.incremental}

- What is a **statistical model**?
- What is a **likelihood**?
- Properties of the **maximum likelihood estimator**

:::

# Statistical models

# Defining a statistical model

::: {.incremental}

- In the regression setting we tried to make as few assumptions about the data-generating process as possible.
  - Our goal is just to estimate and conduct inference on $E[Y|X]$.

- But what if we wanted to make further probabilistic statements about other quantities beyond $\beta$
  - (e.g.) Can we provide a distribution for $Y_{n+1}$, the "next" observation given $X_{n+1}$?
  - If we're willing to make more assumptions about the data-generating process, we can do a lot more!

- **Statistical models** specify the data-generating process in terms of *systematic* and *stochastic* components.
  - **Systematic** elements are functions known constants and unknown *parameters*
  - **Stochastic** elements are draws from probability distributions

- We will be primarily working with *parametric* models
  - The data will be assumed to come from a particular family of probability distributions
  - The "structure" of the model is assumed fixed (the number of parameters does not grow with the size of the data).

:::

# The linear model

::: {.incremental}

- It is common to see the linear model written in its fully parametric form.
- **Stochastic**:

  $$Y_i \sim \text{Normal}(\mu_i, \sigma^2)$$

- **Systematic**:

  $$\mu_i = X_i^{\prime}\beta$$

- What's assumed to be known?
  - $\mathbf{X}$
- What's assumed to have a particular distribution?
  - $Y$
- We are interested in estimating and conducting inference on the parameters: $\beta$ and (less importantly) $\sigma^2$.

:::

# General model notation

::: {.incremental}

- We can specify a broad set of models for $Y_i$ using this framework
- **Stochastic**

  $$Y_i \sim f(\theta_i, \alpha)$$

- **Systematic**

  $$\theta_i = g(X_i, \beta)$$

- What are these quantities?
  - $Y_i$ is a random variable
  - $f()$ denotes the distribution of that random variable
  - $\theta_i$ and $\alpha$ are parameters of that distribution
  - $g()$ is some function 
  - $X_i$ are observed, known constants (e.g. regressors)
  - $\beta$ are parameters of interest 

- We will spend some time with a particular class of models called "Generalized Linear Models" where the systematic component has the form

  $$\theta_i = g(X_i^{\prime}\beta)$$

:::

# Types of distributions

::: {.incremental}

- **Normal**

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
norm_curve <- ggplot() + xlim(-5, 5) + geom_function(fun=dnorm) + theme_bw() + ylab("Density") + xlab("Value")
norm_curve
```

- Continuous on an unbounded support $(-\infty, \infty)$
- Two parameters: Mean $\mu$ and Variance $\sigma^2$

- Probability Density Function (PDF)

$$f_N(x;\mu, \sigma) = \frac{1}{\sigma\sqrt{2\pi}} \exp\left\{- \frac{1}{2}\left(\frac{x - \mu}{\sigma}\right)^2 \right\}$$

:::

# Types of distributions

::: {.incremental}

- **Poisson**

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
pois_curve <- data.frame(x=0:10) %>% ggplot(aes(x=x)) + xlim(0, 10) + geom_point(aes(y=dpois(x, lambda=1))) + theme_bw() + ylab("Probability") + xlab("Value")
pois_curve
```


- Discrete, defined on the support of the natural numbers (positive integers + zero)
- Single parameter: Mean and Variance $\lambda$ 


- Probability Density Function (PDF)

$$f(x; \lambda) = \frac{\lambda^x\exp\{-\lambda\}}{x!}$$

:::

# Types of distributions

::: {.incremental}

- **Binomial**

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
binom_curve <- data.frame(x=0:10) %>% ggplot(aes(x=x)) + xlim(0, 10) + geom_point(aes(y=dbinom(x, size=10, prob = .3))) + theme_bw() + ylab("Probability") + xlab("Value")
binom_curve
```

- Discrete, defined on the support of integers from $\{0, 1, 2, \dotsc, n\}$ (model the sum of repeated i.i.d. coin flips.)
- Two parameters: $p$ probability of success in $n$ trials
  - Special case where $n=1$ trials is typically called the "Bernoulli"
- Probability Density Function (PDF)

$$f(x; p, n) = {n \choose x} p^x(1-p)^{n-x}$$

:::

# Likelihood inference

# Learning about the unknown

::: {.incremental}

- Suppose that I want to learn about an unobserved parameter from a sample of observations.
  - I *assume* a particular statistical model for the data
  - **Example**: The linear model from earlier!

- **Frequentist** approach
  - Unobserved parameters are **fixed constants**
  - Data are **random variables**

- We want to construct an **estimator** that is a *function* of the data and which has desirable properties
  - **Consistency**: As our sample size gets larger, the estimator converges (in probability) to the target parameter.
  - **Asymptotic normality**: In large samples, the distribution of our estimator is normal (ideally with a variance we can estimate as well!)

- Can we come up with a *generic* framework that yields a "good" estimator for a large class of statistical models?

:::

# The Likelihood Function

::: {.incremental}

- Consider data $\mathbf{y} = \{y_1, y_2, \dotsc, y_n\}$ which we assume comes from a known distribution with density $f(\mathbf{y} | \theta)$ with unknown parameter $\theta$
  - $f()$ could be Bernoulli, Normal, Poisson, Gamma, Beta, etc... -- in this setting it is a *known* distribution
  - $\theta$ is an *unknown* parameter that lies in some space $\Theta$ of possible values.

- The **likelihood function** is a function of $\theta$ that is evaluated at the observed values of $\mathbf{y}$

  $$\mathcal{L}(\theta| \mathbf{y}) = f(\mathbf{y} | \theta)$$

- Given some *input* $\theta$, the likelihood function returns the density of the observed data evaluated at that value.

- The likelihood function is **not** a probability density
  - a pdf is a function that takes $x$ as an input
  - a likelihood is a function that takes $\theta$ as an input
  
:::

# The Likelihood Function

::: {.incremental}

- The likelihood function is **not** $f(\theta | \mathbf{y})$
  - That statement doesn't even make sense in a frequentist framework - parameters are constant

- Even when we (later) move to a *Bayesian* framework $f(\theta | \mathbf{y}) \neq f(\mathbf{y} | \theta)$

- Remember Bayes' Rule:
  
  $$f(\theta | \mathbf{y}) = \frac{f(\mathbf{y} | \theta)}{f(\mathbf{y})} \times f(\theta)$$

- Or alternatively, since $f(\mathbf{y})$ is just a normalizing constant, you'll see this written as:
  
  $$\underbrace{f(\theta | \mathbf{y})}_{\text{posterior}} \propto \underbrace{f(\mathbf{y}|\theta)}_{\text{likelihood}} \times \underbrace{f(\theta)}_{\text{prior}}$$

:::

# The Likelihood Function

::: {.incremental}

- We often make the assumption that our data are **independently and identically distributed**
  - $y_i \underset{{\text{i.i.d}}}{\sim} f(y_i | \theta)$

- This allows us to factor the likelihood 

  $$\mathcal{L}(\theta|\mathbf{y}) = f(\mathbf{y} | \theta) = \prod_{i=1}^n f(y_i | \theta)$$

- Since our eventual goal will be to find an *optimum* of the likelihood, we can apply any monotonic function to it since that preserves maxima/minima.
  - The main one we'll apply is the **logarithm**
  - Why? Because logs turn annoying-to-work-with products into easier-to-work-with sums!

- The log-likelihood is typically denoted $\ell(\theta| \mathbf{y})$

  $$\ell(\theta|\mathbf{y}) = \log f(\mathbf{y}|\theta) = \sum_{i=1}^n \log f(y_i | \theta)$$

:::

# MLE

::: {.incremental}

- We want to come up with an estimator $\hat{\theta}$ for the parameter $\theta$
  - $\hat{\theta}$ is a function of the data (like a sample mean or OLS coefficient)
  - Is there a principled way to pick $\hat{\theta}$ that has provably "good" properties across a wide variety of models?

- The Maximum Likelihood Estimator (MLE) $\hat{\theta}$ is defined as the value of $\theta$ that yields the optimum value of the likelihood function

$$\hat{\theta} = \argmax_\theta \ \log f(\mathbf{y}|\theta)$$

- Under i.i.d. observations, the MLE has some desirable properties:
  - The MLE $\hat{\theta}$ is consistent for the true parameter $\theta$
  - It's asymptotically normal: $\sqrt{n}(\hat{\theta} - \theta) \rightarrow \text{Normal}(0, \frac{1}{\mathcal{I}(\theta)})$
  - It's asymptotically efficient: It's variance approaches the Cramer-Rao lower-bound $\frac{1}{n\mathcal{I}(\theta)}$

:::

# Score

::: {.incremental}

- Before illustrating consistency, we need to define some additional functions of the log-likelihood.
- The first is the **score** or the gradient of $\ell(\theta|\mathbf{y})$ with respect to $\theta$

  $$\ell^\prime(\theta|\mathbf{y}) = \frac{\partial}{\partial\theta} \log f(\mathbf{y}|\theta)$$

- With i.i.d. data, you'll often see the likelihood and score written as a mean over the individual likelihoods

  $$\ell(\theta|\mathbf{y}) = \frac{1}{n}\sum_{i=1}^n \log f(y_i|\theta)$$

  $$\ell^\prime(\theta|\mathbf{y}) = \frac{1}{n}\sum_{i=1}^n\frac{\partial}{\partial\theta} \log f(y_i|\theta)$$


- At the value of the true parameter $\theta_0$, the **expected value** of the score is $0$ (under some regularity conditions)

  $$\mathbb{E}\bigg[\frac{\partial}{\partial\theta} \log f(y_i|\theta_0)\bigg] = \frac{\partial}{\partial\theta} \mathbb{E}[\log f(y_i|\theta_0)] = 0$$
  
:::

# Illustration: The Normal Model

::: {.incremental}

- Let's consider the classical **normal regression** model with homoskedastic errors

   $$y_i \sim \text{Normal}(\mu_i, \sigma^2)$$
   $$\mu_i = x_i^{\prime}\beta$$
- Write down the log-likelihood

  $$\ell(\beta, \sigma^2 | \mathbf{y}) = \log\bigg(f(\mathbf{y} | \beta, \sigma^2)\bigg)$$
- Under i.i.d. the joint density of $\mathbf{Y}$ factors into the marginal densities

  $$\ell(\beta, \sigma^2 | \mathbf{y}) = \log\bigg(\prod_{i=1}^n f(y_i| \beta, \sigma^2)\bigg)$$

:::

# Illustration: The Normal Model

::: {.incremental}

- Properties of logs
  
  $$\ell(\beta, \sigma^2 | \mathbf{y}) = \sum_{i=1}^n \log f(y_i| \beta, \sigma^2)$$

- Plug in the Normal density for each observation

  $$\ell(\beta, \sigma^2 | \mathbf{y}) = \sum_{i=1}^n \log \bigg[\frac{1}{\sigma\sqrt{2\pi}} \exp\left\{- \frac{1}{2}\left(\frac{y_i - x_i^\prime\beta}{\sigma}\right)^2 \right\}\bigg]$$
  
- Properties of logs

  $$\ell(\beta, \sigma^2 | \mathbf{y}) = \sum_{i=1}^n -\log(\sigma) - \log(2\pi) + \log\bigg[\exp\bigg\{-\frac{1}{2}\bigg(\frac{y_i - x_i^\prime\beta}{\sigma}\bigg)^2\bigg\}\bigg]$$
  
:::

# Illustration: The Normal Model

::: {.incremental}

- Drop additive constants; log is the inverse of exp

  $$\ell(\beta, \sigma^2 | \mathbf{y}) = \sum_{i=1}^n -\log(\sigma) -\frac{1}{2}\bigg(\frac{y_i - x_i^\prime\beta}{\sigma}\bigg)^2$$
  
- Algebra

  $$\ell(\beta, \sigma^2 | \mathbf{y}) =  -\frac{1}{2\sigma^2} \sum_{i=1}^n (y_i - x_i^\prime\beta)^2 -n\log(\sigma)$$

- What's familiar here?
  - The objective function is the sum of squared residuals!
  - We know the solution already - it's OLS.

:::
  
# Consistency

::: {.incremental}

- With this definition of the score, we can show consistency of the MLE under i.i.d. observations.
- By the weak law of large numbers

  $$\frac{1}{n} \sum_{i=1}^n \log f(y_i | \theta) \underset{p}{\rightarrow} \mathbb{E}[\log f(y_i | \theta)]$$

- The MLE is an optimizer of the left-hand side.
  - Therefore it converges in probability to the optimizer of the right-hand side

- And we just showed that the expectation of the score is $0$ at the true value of $\theta$, denoted $\theta_0$
  - Therefore $\theta_0$ is the optimizer of the right-hand side

- Therefore the MLE $\hat{\theta}$ is consistent for the true value $\theta_0$

:::

# Consistency

::: {.incremental}

- Note a few important conditions for identification

- **Identifiability**
  - No "plateaus" in the log-likelihood.
  - Two different values of $\theta$ can't both maximize the log-likelihood
  
  $$f(\mathbf{y}|\theta) \neq f(\mathbf{y} | \theta_0) \ \forall \ \theta \neq \theta_0$$ 
  
- **Fixed parameter space**
  - The dimensionality of $\theta$ stays fixed and does not depend on $n$
  - Violated (e.g.) in ideal point models (new legislators mean new ideal points)

- $\theta_0$ is in the **interior** of the parameter space

:::

# Information

::: {.incremental}

- Having established consistency, we're interested in understanding the variance and distribution (asymptotically) of $\hat{\theta}$ 
  - To do this, we need to define a quantity called the **information**: $\mathcal{I}_n$
  - This is equivalent to the variance of the score at its optimum

  $$\mathcal{I}_n = \mathbb{E}\left[\left(\frac{\partial}{\partial\theta} \log f(\mathbf{y}|\theta_0)\right)^2\right]$$

- We can show (under some more regularity conditions) that this can be written as the negative expectation of the second-order partial derivative (Hessian) of the log-likelihood

  $$\mathcal{I}_n = -\mathbb{E}\left[\frac{\partial^2}{\partial\theta^2} \log f(\mathbf{y}|\theta_0)\right]$$

- **Intuitively**: Captures the curvature of the log-likelihood at its maximum

:::


# Information

- You'll sometimes see the information written in terms of a single observation

  $$\mathcal{I} =  -\mathbb{E}\left[\frac{\partial^2}{\partial\theta^2} \log f(y_i|\theta_0)\right]$$

- Under our i.i.d. assumption, we can write

  $$\mathcal{I_n} = n\mathcal{I}$$

- This matters for how we write/show consistency of the MLE

# Asymptotic distribution

- Next, we can show that in large samples, the MLE under i.i.d. observations converges to a normal distribution with a known variance (the inverse Fisher information)

 $$\sqrt{n}(\hat{\theta} - \theta) \overset{d}{\rightarrow} \text{Normal}(0, \mathcal{I}^{-1})$$
 
- We'll write the "normalized" likelihood under i.i.d.

  $$\ell(\theta|\mathbf{y}) = \frac{1}{n}\sum_{i=1}^n \log f_i(y_i | \theta)$$

- And the score

  $$\ell^{\prime}(\theta|\mathbf{y}) = \frac{1}{n}\sum_{i=1}^n \frac{\partial}{\partial \theta} \log f_i(Y_i | \theta)$$

- And the Hessian

  $$\ell^{\prime\prime}(\theta|\mathbf{y}) = \frac{1}{n} \sum_{i=1}^n \frac{\partial^{2}}{\partial \theta^{2}}  \log f_i(Y_i | \theta)$$

# Asymptotic distribution

- Our first step will be to take a first-order **taylor approximation** of the score function around the true MLE $\theta_0$ and write it as a function of our estimator $\hat{\theta}$

  $$\ell^{\prime}(\hat{\theta}|\mathbf{y}) = \ell^{\prime}(\theta_0 | \mathbf{y}) + \ell^{\prime\prime}(\theta_0|\mathbf{y})(\hat{\theta} - \theta_0)$$
  
- Using the fact that the score is zero at the MLE, we have

  $$0 = \ell^{\prime}(\theta_0 | \mathbf{y}) + \ell^{\prime\prime}(\theta_0|\mathbf{y})(\hat{\theta} - \theta_0)$$
- Rearranging gives

$$(\hat{\theta} - \theta_0) = [-\ell^{\prime\prime}(\theta_0|\mathbf{y})]^{-1}\ell^{\prime}(\theta_0|\mathbf{y})$$

- Multiplying by $\sqrt{n}$

$$\sqrt{n}(\hat{\theta} - \theta_0) = [-\ell^{\prime\prime}(\theta_0|\mathbf{y})]^{-1}[\sqrt{n}\ell^{\prime}(\theta_0|\mathbf{y})]$$

# Asymptotic distribution

- First, by law of large numbers, 

  $$-\ell^{\prime\prime}(\theta_0|\mathbf{y}) = -\frac{1}{n} \sum_{i=1}^n \frac{\partial^2}{\partial\theta^2} \log f(y_i|\theta_0) \overset{p}{\to}  -\mathbb{E}\left[\frac{\partial^2}{\partial\theta^2} \log f(y_i|\theta_0)\right] = \mathcal{I}$$
  
- Second, by the central limit theorem

  $$\sqrt{n} \ell^{\prime}(\theta_0 | \mathbf{y}) = \frac{1}{\sqrt{n}} \sum_{i=1}^n \frac{\partial}{\partial\theta} \log f(y_i | \theta_0) \overset{d}{\to} \mathcal{N}\bigg(0, Var\bigg( \frac{\partial}{\partial\theta} \log f(y_i | \theta_0)\bigg)\bigg) = \mathcal{N}(0, \mathcal{I})$$
  
- Lastly, by Slutsky's theorem (limit of a function is a function of the limits)

  $$\sqrt{n}(\hat{\theta} - \theta_0) \overset{d}{\to} \mathcal{N}(0, \mathcal{I}^{-1})$$
  
# Asymptotic distribution

- So in large samples, the MLE is approximately normally distributed, centered at the true parameter $\theta_0$ and with a variance equal to the **inverse** of the information

  $$\hat{\theta} \sim \mathcal{N}(\theta_0, \mathcal{I}_n^{-1})$$

- Can another estimator do better asymptotically?
  - It turns out the answer is **no** 
  - The MLE is also asymptotically the most efficient unbiased estimator since its asymptotic variance is equal to the **Cramér–Rao lower bound**

# Cramér–Rao lower bound (CRLB)

- An important result in statistics relates to the smallest possible variance that an **unbiased** estimator could reach
  - That is, if we're not willing to accept more **bias**, how much **precision** can we squeeze out of the data?
  - This ends up being related to the information!

- The **Cramér–Rao lower bound** on the variance of any unbiased estimator of $\theta$, $\hat{\theta}$ is

  $$Var(\hat{\theta}) \ge (\mathcal{I}_n)^{-1}$$
  
  
# Cramér–Rao lower bound (CRLB)

- Consider what **unbiasedness** gets us:

  $$\mathbb{E}[\hat{\theta} - \theta] = \int(\hat{\theta} - \theta) f(\mathbf{y}| \theta)d\mathbf{y} = 0$$
  
- Since unbiasedness is true for every value of $\theta$, differentiating with respect to $\theta$ should also give us 0

  $$\frac{\partial}{\partial\theta} \int(\hat{\theta} - \theta) f(\mathbf{y} |\theta)d\mathbf{y} = 0$$
  
- Under the conditions that let us exchange differentiation and integration
  
  $$ \int \frac{\partial}{\partial\theta} \bigg((\hat{\theta} - \theta) f(\mathbf{y}| \theta)\bigg)d\mathbf{y} = 0$$
  
# Cramér–Rao lower bound (CRLB)

- Product Rule
  
  $$\int  \bigg((\hat{\theta} - \theta) \frac{\partial}{\partial\theta} f(\mathbf{y}| \theta)\bigg)d\mathbf{y} - \int  f(\mathbf{y}| \theta)d\mathbf{y} = 0$$

- PDFs integrate to 1 over their support

  $$\int  \bigg((\hat{\theta} - \theta) \frac{\partial}{\partial\theta} f(\mathbf{y}| \theta)\bigg)d\mathbf{y} = 1$$
  
- We can write $\frac{\partial}{\partial \theta} f(\mathbf{y}| \theta)$ in terms of the **log** density (via chain rule)

  $$\frac{\partial}{\partial \theta} f(\mathbf{y}; \theta) = f(\mathbf{y}| \theta) \frac{\partial}{\partial \theta} \log  f(\mathbf{y}|\theta)$$
  
# Cramér–Rao lower bound (CRLB)

- Plugging back in!

  $$\int  \bigg((\hat{\theta} - \theta)  \frac{\partial}{\partial \theta} \log  f(\mathbf{y}|\theta)\bigg)f(\mathbf{y}| \theta)d\mathbf{y} = 1$$

- "Law of the unconscious statistician" (definition of expectations of functions)

$$\mathbb{E}\bigg[(\hat{\theta} - \theta)  \times \frac{\partial}{\partial \theta} \log  f(\mathbf{y}|\theta)\bigg] = 1$$

- Square both sides

$$\bigg|\mathbb{E}\bigg[(\hat{\theta} - \theta)  \times \frac{\partial}{\partial \theta} \log  f(\mathbf{y}|\theta)\bigg]\bigg|^2 = 1$$


# Cramér–Rao lower bound (CRLB)

- (One formulation of) the **Cauchy-Schwarz Inequality** for expectations states that for two random variables $X$ and $Y$

  $$|\mathbb{E}[XY]|^2 \le \mathbb{E}(X^2)\mathbb{E}(Y^2)$$
  
- Therefore, by **Cauchy-Schwarz**

  $$\mathbb{E}[(\hat{\theta} - \theta)^2] \times \mathbb{E}\bigg[\bigg(\frac{\partial}{\partial \theta} \log f(\mathbf{y}|\theta)\bigg)^2\bigg] \ge 1$$
  
- First term is the variance of an unbiased estimator (MSE). Second term is the information. Rearranging gives us the CRLB

  $$Var(\hat{\theta}) \ge \frac{1}{\mathcal{I}_n}$$

# Asymptotic normality

- Under our i.i.d. assumption (+ regularity conditions), the Central Limit Theorem lets us show that the MLE is asymptotically normal

  $$\sqrt{n}(\hat{\theta} - \theta) \overset{d}{\rightarrow} \text{Normal}(0, \mathcal{I}^{-1})$$
  
- The key intuition is again from the fact that we can write the 


- So in large samples, a good approximation of the variance of $\hat{\theta}$ is $\mathcal{I}_n^{-1}$
  - An additional result: The inverse of the information is the lowest possible variance of any unbiased estimator (**Cramér-Rao lower bound**) 
  - MLE achieves this bound asymptotically


- How do we do inference then? Same way as before -- plug in a consistent estimator of $\mathcal{I}_n^{-1}$.
  - Calculate the hessian of the log-likelihood at the MLE and take the inverse of its negative.

# Illustration: The Normal Model

# Conclusion

- **Statistical models**
  - Describe the data-generating process in terms of *systematic* and *stochastic* components
  - In a parametric model, we assume the data $\mathbf{y}$ come from a known *distribution* with unknown parameters $\theta$

- **Likelihood**
  - A function of $\theta$ evaluated at the observed data $\mathbf{y}$ equal to $f(\mathbf{y}|\theta)$
  - How "likely" are my observed results in a world where the true DGP parameter were $\theta$.
  - Not $f(\theta|\mathbf{y})$. That quantity only makes sense in the Bayesian context and requires us to formulate a prior $f(\theta)$

- **Maximum Likelihood Estimator**
  - A *technique* to come up with a **good** estimator in any case where we can write down a (well-behaved) likelihood function for a DGP
  - Still a "function" of the data, but that function has useful properties:
    - Consistency, Asymptotic Normality, Asymptotic Efficiency
    
---
