---
title: "Week 12: Flexible Regression - Part 3"
subtitle: "PS 818 - Statistical Models"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "November 17, 2025"
date-format: "MMMM D, YYYY"
editor: 
  render-on-save: true
format: 
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 818 - University of Wisconsin - Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute: 
  cache: true
code:
  execute: 
    - r
    - python
embed-resources: true
---


$$
  \require{cancel}
  \DeclareMathOperator*{\argmin}{arg\,min}
$$


$$
 \DeclareMathOperator*{\argmax}{arg\,max}
$$


```{r, echo = F, message=F, warnings=F}
library(tidyverse)
library(haven)
library(estimatr)
library(knitr)
library(splines)
library(lubridate)

options(digits=3)
knitr::opts_chunk$set(echo = T)
set.seed(60639)

library(caret)
library(rpart)

```


# Previously

$$
  \DeclareMathOperator*{\argmax}{arg\,max}
$$

::: {.incremental}

- **Kernels**
  - "Kernel smoothing" - Kernels as a "localization" tool
  - "Kernel methods" - Kernels as regression inputs
  - Two *different* uses of the term/concept - but some interesting relations
- **Forests**
  - Tree-based models for classification and regression
  - Ensembles/averages of trees via **random forests**

:::

# This week

::: {.incremental}

- **More forests**
  - Other methods of aggregating trees - **Bayesian Additive Regression Trees**
  - Trees for detecting causal heterogeneity
  - "Honest" inference via sample-splitting

- **More kernels**
  - Model-based interpretation via **gaussian processes**

:::

# Classification and Regression Trees (CART)

::: {.incremental}

- The **CART** algorithm constructs a **prediction** for an outcome $Y$ given some inputs $X$ by **partitioning** the space of $X$ into a series of $M$ **hyper-rectangles**
  - Introduced by **Breiman et. al. (1984)**
  - Very good political science summary in **Montgomery and Olivella (2018)**
  
- We'll define a **tree**-based prediction as:

$$f(X) = T(X, \theta) = \sum_{m=1}^M c_m \mathbf{I}(X \in R_m)$$

- Here $\theta$ denotes the tree-specific "hyperparameters" (e.g. tree-depth, node size, etc...)

:::

# Classification and Regression Trees (CART)

::: {.incremental}

- Searching over the space of all possible splits and all possible covariates to minimize some error function is **computationally prohibitive**
- Basic **recursive binary partitioning** algorithm:
  - (1): For each variable, find a splitting point that minimizes the **within-partition** variance
  - (2): Split on the single variable that provides the **best split** among these variables
  - (3): Repeat 1/2 on each **sub-sample** created by the previous split.
  - (4): Stop if some stopping rule reached (e.g. minimum node size).

- **Prune** the tree to remove splits that have a minimal impact on fit

:::


# Example: Transgender attitudes (Broockman and Kalla, 2016)

- **Broockman and Kalla (2016)** study the effect of in-person canvassing on attitudes towards transgender individuals.
  - We'll get to the experiment later, but first we'll model attitudes at baseline using available covariates.

- The study started by recruiting about 68,000 respondents for a baseline intake survey

```{r}
canvas <- read_dta("data/broockman_kalla_replication_data.dta")
```

- It then randomly selected 1,825 to be assigned to receive either of the two canvassing conditions.

```{r}
canvas_exp <- canvas %>% filter(!is.na(treat_ind))
canvas_exp <- canvas %>% filter(!is.na(vf_age)&!is.na(vf_republican)&!is.na(vf_democrat)&!is.na(vf_female)&!is.na(vf_hispanic)&
                        !is.na(religious_t0)&!is.na(ssm_t0)&!is.na(ideology_t0))
```

# Example: Transgender attitudes 

- The main outcome of interest is `therm_trans_t0`, a 0-100 attitude thermometer measured at baseline.

```{r, echo=F, fig.align='center', fig.dpi=100}
canvas_exp %>% ggplot(aes(x=therm_trans_t0)) + geom_histogram(bins=35) + theme_bw() + labs(x="Transgender Feeling Thermometer (0-100)", y="Count")
```

# Example: Transgender attitudes

- We have some covariates observed in the voter file as well as a number of questions asked on the survey

```{r}
predict_tree <- rpart(therm_trans_t0 ~ vf_age + vf_republican + vf_democrat + vf_female + vf_hispanic +
                        religious_t0 + ssm_t0 + ideology_t0,
                  data = canvas_exp)
```

# Example: Transgender attitudes

```{r, message = F, warning=F, fig.width=11, fig.height=6, fig.align="center", echo=F, dpi=150}
rattle::fancyRpartPlot(predict_tree, sub="")
```

# Example: Transgender attitudes

- What happens if we lower the "complexity parameter" required to even consider a split

```{r}
predict_tree_moresplits <- rpart(therm_trans_t0 ~ vf_age + vf_republican + vf_democrat + vf_female + vf_hispanic +
                        religious_t0 + ssm_t0 + ideology_t0,
                  data = canvas_exp, control = rpart.control(cp=0.005))
```

```{r}
plotcp(predict_tree_moresplits)
```

# Example: Transgender attitudes

```{r, message = F, warning=F, fig.width=11, fig.height=6, fig.align="center", echo=F, dpi=150}
rattle::fancyRpartPlot(predict_tree_moresplits, sub="")
```


# Example: Transgender attitudes

- Turns out attitudes on same-sex marriage are **extremely** predictive. What if we took out that input?

```{r}
predict_tree2 <- rpart(therm_trans_t0 ~ vf_age + vf_republican + vf_democrat + vf_female + vf_hispanic +
                        religious_t0 + ideology_t0,
                  data = canvas_exp)
```

# Example: Transgender attitudes

```{r, message = F, warning=F, fig.width=11, fig.height=6, fig.align="center", echo=F, dpi=150}
rattle::fancyRpartPlot(predict_tree2, sub="")
```

# Example: Transgender Attitudes

- Lastly, what about just the voter file covariates

```{r}
predict_tree3 <- rpart(therm_trans_t0 ~ vf_age + vf_republican + vf_democrat + vf_female + vf_hispanic,
                  data = canvas_exp)
```

# Example: Transgender attitudes

```{r, message = F, warning=F, fig.width=11, fig.height=6, fig.align="center", echo=F, dpi=150}
rattle::fancyRpartPlot(predict_tree3, sub="")
```

# Combining trees

::: {.incremental}

- The problem with a **single tree** is that it's still highly likely to overfit the data even with pruning and cross-validation.

- **Solution**: Grow many trees and combine them

  $$f(X) = \sum_{k=1}^K T_k(X, \theta_k)$$

- But lots of different approaches to **combination**:
  - **Random forests**
  - **Gradient Boosting**
  - **Bayesian Additive Regression Trees**
  
:::

# Random forests

::: {.incremental}

- The **easiest** way to aggregate is to simply average over a bunch of trees grown in different ways
  - Best aggregation comes from having **independent** trees
  
- **Breiman (2001)** Random Forest algorithm **de-correlates** trees through two techniques
  - **Bootstrap aggregatting** (Bagging) - Growing trees on bootstrap re-samples of the data
  - **Random subspaces**/"Feature bagging" - Growing trees on randomly sampled subsets of the features
  
- Aggregate using simple averages of predictions from each tree

:::

# Gradient boosting

::: {.incremental}

- Alternatively, we can think of **iteratively** growing trees that predict the residuals from the combination of all **previous** models
  - "Boosting" transforms a sequence of "weak" learners (single trees) into a collectively stronger predictor
  
- **Example**: In our previous illustration, we found that attitudes towards **same sex marriage** were the strongest predictor in the first tree
  - What happens if we fit a tree to the **residuals** from the first tree?
  
```{r}
canvas_exp$therm_trans_t0_resid1 <- canvas_exp$therm_trans_t0 - predict(predict_tree_moresplits)
predict_tree_residual <- rpart(therm_trans_t0_resid1 ~ vf_age + vf_republican + vf_democrat + vf_female + vf_hispanic +
                        religious_t0 + ssm_t0 + ideology_t0, control = rpart.control(cp=0.005),
                  data = canvas_exp)
```
  
:::

# Gradient boosting

```{r, message = F, warning=F, fig.width=11, fig.height=6, fig.align="center", echo=F, dpi=150}
rattle::fancyRpartPlot(predict_tree_residual, sub="")
```

# Gradient boosting

- Each tree is grown to explain the **residual** left behind by the combination of **all previous** trees.
  - New trees are "shrunk" by some factor between $0$ and $1$

```{r}
library(gbm3) #install_github("gbm-developers/gbm3")
boosted_trees <- gbmt(therm_trans_t0 ~ vf_age + vf_republican + vf_democrat + vf_female + vf_hispanic +
                        religious_t0 + ssm_t0 + ideology_t0,  data = canvas_exp)
```

# Variable importance

```{r, message = F, warning=F, fig.width=11, fig.height=6, fig.align="center", echo=F, dpi=150}
summary(boosted_trees)
```

# Partial dependence plot

```{r, message = F, warning=F, fig.width=11, fig.height=6, fig.align="center", echo=F, dpi=150}
plot(boosted_trees, var_index = c(1,7))
```

# Bayesian Additive Regression Trees (BART)

- **BART** is also a "sum-of-trees" model
  - Aggregate many individual trees into a "strong" predictor of $Y$
  
- But regularizes trees using a fully Bayesian outcome model

$$Y_i = \sum_{k=1}^K T_k(X_i, \theta_k) + \epsilon_i$$
$$\epsilon_{i} \sim \text{Normal}(0, \sigma^2)$$

- Place a Bayesian prior over both the tree hyperparameters $\theta_k$ (e.g. tree depth), predictions at each node ($\mu$) and outcome variance $\sigma^2$
  - Estimation via MCMC

# Causal Trees

::: {.incremental}

- A common task in social science is not **prediction** but rather **description** of **heterogeneity**
  - Researchers run **randomized experiments** to identify **average** treatment effects.
  
- But we often want to know how the treatment effect **varies** by individuals - we want **conditional average treatment effects**
  - For which units is the effect **positive**?
  - For which units is the effect **negative**?
  
- Because of the **fundamental problem of causal inference**, we can't observe individual effects directly
  - Sometimes we have a clear idea of which groups we'd like to test heterogeneity for **ex-ante**
  - But **ex-post** we're worried about the **garden of forking paths**.
  
- Can we find **heterogenous treatment effects** without knowing the sub-groups in advance and without p-hacking?

:::

# Causal Trees

::: {.incremental}

- **Athey and Imbens (2016)** propose using trees to model heterogeneity.
  - Find a **partition** of the sample into subsets with **distinctive** treatment effects
  
- Modifies the off-the-shelf CART in two key ways:
  1. Optimizes splits for **treatment effects** (where we can't observe the 'ground truth')
  2. "Honest" inference via sample splitting
  
- **Key intuitions**
  1. Splits for effect heterogeneity + precisely estimated treatment effects at each leaf
  2. Avoids over-fitting by estimating effects on a different sample than the one used to create the splits
  
:::

# Honest inference

::: {.incremental}

- We might be concerned about **over-splitting** our sample to random noise in the treatment effects.
  - The sample split criterion alone is always guaranteed to find the split that **maximizes** the difference in effects across nodes.
  
- **Solution** - Evaluate the treatment effects using a **separate sample**
  - "Honest" inference splits the sample randomly in two
  - The first sample is used to generate the splits for a tree
  - The second sample is used to **estimate the effects** at each node.

- Guarantees we still have unbiased inference on the treatment effects **given** a partition fit to optimize effect heterogeneity. 

:::

# Example: Effect of Canvassing

- Let's go back to **Broockman and Kalla (2016)** 
  - What's the effect of the canvassing treatment at the first time period?
  
```{r}
canvas_exp_t1 <- canvas_exp %>% filter(!is.na(therm_trans_t1))
canvas_exp_t1 <- canvas_exp_t1 %>% mutate(therm_trans_chg = therm_trans_t1 - therm_trans_t0)
ate <- lm_robust(therm_trans_chg ~ treat_ind, data=canvas_exp_t1)
summary(ate)
```

- Does this vary based on our pre-treatment covariates?

# Example: Effect of Canvassing

- If we just want a **single tree**, we can use `htetree`
  - Need to **split** our sample into training and evaluation
  
```{r}
library(htetree)
set.seed(53)
indices <- sample(1:nrow(canvas_exp_t1), nrow(canvas_exp_t1)/2)
train_canvas_t1 <- canvas_exp_t1[indices,]
eval_canvas_t1 <- canvas_exp_t1[-indices,]

trans_tree <- honest.causalTree(therm_trans_chg ~ vf_age + vf_republican + vf_democrat + vf_female + vf_hispanic +
                        religious_t0 + ssm_t0 + ideology_t0, data = train_canvas_t1,
     treatment = train_canvas_t1$treat_ind,
     est_data = eval_canvas_t1,
     est_treatment = eval_canvas_t1$treat_ind,
     split.Rule = "CT",
     cv.option = "CT", split.Honest = T, cv.Honest = T, split.Bucket = F, 
                   xval = 10, cp = 0, minsize = 20, propensity = 0.5)
```

# Example: Effect of Canvassing

```{r, message = F, warning=F, fig.width=11, fig.height=6, fig.align="center", echo=F, dpi=150}
rattle::fancyRpartPlot(trans_tree, sub="")
```

# Causal forests

::: {.incremental}

- A **single** tree might still overfit and fail to accurately capture the true effect heterogeneity.
  - Unstable splits!
  - The presence of **sharp** decision boundaries means we're unlikely to have a good estimator for the CATE $\hat{\tau}(X_i)$

- **Solution**: Do forests (**Athey and Wager (2018)**)
  - Plug in the Causal Tree algorithm into the standard random forest methodology (bagging + random subspaces)
  
- **Downsides**:
  - No longer have an **interpretable** set of sub-groups.
  - However, see **Huang, Tang and Kenney (2025)** for a new method for recovering these from a tree fit to the random forest predicted $\hat{\tau}(X_i)$

:::

# Example: Effect of Canvassing

- `grf` ("generalized random forests") implements this approach

```{r}
library(grf)
set.seed(53703)
trans_forest <- causal_forest(X =model.matrix( ~ vf_age + vf_republican + vf_democrat + vf_female + vf_hispanic +
                        religious_t0 + ssm_t0 + ideology_t0,  data = canvas_exp_t1), 
                        Y = as.numeric(canvas_exp_t1$therm_trans_chg), 
                        W = as.numeric(canvas_exp_t1$treat_ind))
print(trans_forest)
```

# Example: Effect of Canvassing

- Let's see what the distribution of treatment effects is in-sample

```{r}
hist(predict(trans_forest)$predictions, xlab="Predicted CATE", ylab="Count", main = "Causal Forest CATEs")
```

# Example: Effect of Canvassing

- We can test out the **causal distillation** method to essentially fit a conventional CART tree to the predicted effects from the causal forest!

```{r}
library(causalDT)
set.seed(53703)
trans_distill <- causalDT(
X =model.matrix( ~ vf_age + vf_republican + vf_democrat + vf_female + vf_hispanic +
                        religious_t0 + ssm_t0 + ideology_t0,  data = canvas_exp_t1), 
                        Y = as.numeric(canvas_exp_t1$therm_trans_chg), 
                        Z = as.numeric(canvas_exp_t1$treat_ind),
holdout_prop = 0.5,
teacher_model = "causal_forest",
student_model = "rpart",
rpart_control = list(cp = 0, minbucket = 20),
rpart_prune = "min"
)
```

# Distilled Effect Heterogeneity

```{r, message = F, warning=F, fig.width=11, fig.height=6, fig.align="center", echo=F, dpi=150}
plot_cdt(trans_distill)
```
  