---
title: "Week 15: End of Class Review"
subtitle: "PS 818 - Statistical Models"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "December 8, 2025"
date-format: "MMMM D, YYYY"
editor: 
  render-on-save: true
format: 
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 818 - University of Wisconsin - Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute: 
  cache: false
code:
  execute: 
    - r
    - python
embed-resources: true

---

# Quantitative Social Science and Political Science

$$
  \require{cancel}
  \DeclareMathOperator*{\argmin}{arg\,min}
$$


$$
 \DeclareMathOperator*{\argmax}{arg\,max}
$$

```{r, echo = F, message=F, warnings=F}
library(tidyverse)
library(haven)
library(estimatr)
library(knitr)
options(digits=3)
```

# The Ideal Methods Sequence

::: {.incremental}

- **Intro to Statistics**
  - Fundamentals of probability theory
  - Principles of statistical inference
  - Point/Interval Estimation; Hypothesis Testing
  - Where does **uncertainty** come from in the research process?

- **Causal Inference**
  - What is a causal effect?
  - What assumptions do we need to learn these from data?
  - What can we do when we can't randomize?
  
- **Statistical Models**
  - What is a statistical model?
  - How do we estimate the **parameters** of the model?
  - How do we generate **predictions**?
  - How do we evaluate the quality of a predictive model?
  - How can we use models to **describe** and **summarize**
  
:::

# My philosophy in a nutshell

::: {.incremental}

1. The goal of statistics in science is to properly characterize error in the process of observation.
2. Models aren't real.
3. Causal effects **are** real.
4. "Mechanisms" are either statements about mediation or interaction
5. Models are useful if they can predict or if they can summarize

:::

# A Typology of Political Science "Quant"

::: {.incremental}

- **Experimentalists**
  - Psychology-based experimentalists
  - Stats/econometrics-based experimentalists

- **Observational Causal Inference**
  - "Credibility revolution"
  - Model-based inference
  
- **Observational** (non-causal)
  - "Structural" estimation 
  - Model-based description

- **Pure description**
  - Public opinion surveys

:::

# A Typology of Political Science "Quant"

::: {.incremental}

- Where does "text-as-data" fit in?
  - Some of this is just "advanced" methods to replace human coders (e.g. LLMs)
  - But the actual **models** usually fall into "model-based description"

- How about the formal modellers?
  - No empirics/motivating qualitative case (IMO, ideal)
  - Structural estimation of model parameters.
  - "Hypotheses" obtained from comparative statics on model parameters $\leadsto$ standard causal inference

- Some will argue that empirics are useless without a connection to some feature of a theoretical model.
  - I'd disagree - causal effects are still "real" insofar as they correspond to an intervention we could do
  - I will grant that any sort of notion of generalization/transportation needs a theory
  
:::

# This course

::: {.incremental}

1. Linear regression to estimate $E[Y|X]$ using a linear combination of the columns of $X$
2. Generalized Linear Models to estimate $E[Y|X]$ using a linear combination of the columns of $X$ and a **transformation** $g()$
3. Maximum Likelihood to obtain CAN estimates of GLM parameters
4. Bayesian inference on parameters - **posterior** = **likelihood** $\times$ **prior**
5. Bayesian inference for GLMs - Incorporating structure through priors
6. Models for summarizing data: mixture models, factor models
7. Models for flexibly estimating $E[Y|X]$

:::

# What we didn't cover

::: {.incremental}

- Method of moments estimation (common in econometrics/structural estimation).
  - **General concept**: Estimators as *solutions to systems of equations* - you saw this with MLE (we set the score to 0)
  
- Neural networks 
  - Defining input-output models via sequences of **layers**
  - Lots of "reinvention" of concepts from stats (with different names)
  - Building blocks look like familiar models we've seen (e.g. logistic regression)

:::

# Review: OLS

::: {.incremental}

- A **linear predictor** is defined as a linear combination of the explanatory variables and coefficients

  $$m(x) = x^\prime\beta = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \dotsc$$
- The **best linear predictor** is the linear predictor that minimizes the **population** squared prediction error

  $$\beta = \argmin_b \ \mathbb{E}[(Y_i - X_i^\prime b)^2]$$
  
- There's a closed form solution in terms of population expectations.

  $$m(X_i) = X_i^\prime\beta = X_i^\prime (\mathbb{E}[X_iX_i^\prime])^{-1}\mathbb{E}[X_iY_i]$$
  
:::

# Review: OLS

::: {.incremental}

- **Ordinary Least Squares** is an estimator applied to paired observations $Y_i$, and $X_i$

  $$\hat{\beta} = \bigg(\frac{1}{n} \sum_{i=1}^n X_iX_i^\prime\bigg)^{-1}\bigg(\frac{1}{n} \sum_{i=1}^n X_iY_i\bigg)$$

- Often written as:

  $$\hat{\beta} = (\mathbf{X}^\prime\mathbf{X})^{-1}(\mathbf{X}^\prime Y)$$

- When is this consistent for the BLP?
  - When we have a sample from the target population (however you define the population)
  - And no perfect collinearity
  
:::

# Review: OLS

::: {.incremental}

- To do inference on the CEF, we assume that it's equal to the BLP!

- **Assumption 1**: Linearity

  $$Y_i = X_i^\prime\beta + \epsilon_i$$

- **Assumption 2**: Strict exogeneity of the errors

  $$\mathbb{E}[\epsilon | \mathbf{X}] = 0$$
  
- What does this get us?
  - Finite-sample properties of OLS (unbiasedness in addition to consistency!)

- When is linearity always **true**?
  - When we have a fully-saturated model!

:::


# Review: OLS

::: {.incremental}

- To do inference on the CEF, we assume that it's equal to the BLP!

- **Assumption 1**: Linearity

  $$Y_i = X_i^\prime\beta + \epsilon_i$$

- **Assumption 2**: Strict exogeneity of the errors

  $$\mathbb{E}[\epsilon | \mathbf{X}] = 0$$
  
- What does this get us?
  - Finite-sample properties of OLS (unbiasedness in addition to consistency!)

- When is linearity always **true**?
  - When we have a fully-saturated model!

:::

# Review: OLS

::: {.incremental}

- With no other assumptions, inference on the sampling variance of $\hat{\beta}$ takes the form of the "sandwich estimator"

  $$\widehat{Var(\hat{\beta})} =  (\mathbf{X}^\prime\mathbf{X})^{-1} (X^\prime \hat{\Sigma} \mathbf{X}) (\mathbf{X}^\prime\mathbf{X})^{-1}$$

- Under i.i.d. but heteroskedastic observations, we'll plug in $\hat{\Sigma} = \text{diag}(\hat{e_1}^2, \hat{e_2}^2, \dotsc \hat{e_n}^2)$
  - "Heteroskedasticity-robust" SEs (HC1, HC2, HC3, etc...)

- If we assume arbitrary correlation across observations **within** cluster but independence **across** clusters, our plug-in for $\hat{\Sigma}$ is a block diagonal matrix with off-diagonals $\hat{e_i}\hat{e_j}$ for $i, j$ in the same cluster.
  - "Clustered standard errors" 
  
:::

# Some useful-to-remember OLS tricks

::: {.incremental}

- Sum of the OLS residuals is zero

  $$\sum_{i=1}^N (Y_i - \hat{Y_i}) = 0$$
  
- We can go from $Y$ to $\hat{Y}$ via the **projection matrix**

  $$\mathbf{P}_{\mathbf{X}} = \mathbf{X}(\mathbf{X}^\prime\mathbf{X})^{-1}\mathbf{X^\prime}$$
- **Frisch-Waugh-Lovell** - Any OLS coefficient $\hat{\beta}_k$ can be written as

  $$\hat{\beta}_k = \frac{\widehat{\text{Cov}}(Y^{\perp \mathbb{X}_{-k}}, X_k^{\perp \mathbb{X}_{-k}})}{\widehat{\text{Var}}( X_k^{\perp \mathbb{X}_{-k}}) }$$
  
:::

# Generalized Linear Models

::: {.incremental}

- When teaching OLS, we focused on estimating the CEF $E[Y|X]$ with linear regression
  - The CEF is an "optimal predictor" for $Y$ in that it achieves lowest mean-squared error.
  
- We want to make predictions for **out of sample** observations
  - Covariate profiles that don't necessarily exist in our sample.
  
- Our linear CEF assumption might create some problems!
  - With a binary outcome, we might get predictions that are negative or greater than 1 - **impossible probabilities**
  
- GLMs model a **function** of the CEF with the linear predictor

  $$g(E[Y|X]) = X\beta$$

:::

# The logistic regression

::: {.incremental}

- The GLM that you are **most likely** to use is the logistic regression.
  - Estimating **propensity scores** for (augmented) inverse probability weighting
  
- **Example**: **Keriakes et. al. (2000)** "Abciximab provides cost-effective survival advantage in high-volume interventional practice"*
  - Abciximab, an anti-clotting drug, is often used during certain types of heart surgery to reduce bleeding risk.
  - Keriakes et. al. (2000) look at 1472 surgeries in Ohio Heart Health Center
  - Abciximab was administered *non-randomly* -- some types of patients more likely to receive the drug than others

:::

# Logistic regression

::: {.incremental}

- The "logit" or "logistic" GLM models the **log-odds** of a binary outcome as a function of the linear predictor $X_i^{\prime}\beta$

  $$Y_i \underset{\text{i.i.d}}{\sim} \text{Bernoulli}(\pi_i)$$

  $$E[Y_i | X_i] = Pr(Y_i = 1 | X_i) = \pi_i$$

  $$\log\bigg(\frac{\pi_i}{1-\pi_i}\bigg) = X_i^{\prime}\beta$$

- Alternatively, this is written in terms of the "inverse-link" function (the logistic function) that relates $\pi_i$ to $g^{-1}(X_i^{\prime}\beta)$

  $$\pi_i = \frac{\exp(X_i^{\prime}\beta)}{1 + \exp(X_i^{\prime}\beta)} = \frac{1}{1 + \exp(-X_i^{\prime}\beta)}$$

:::

# Maximum Likelihood

::: {.incremental}

- We obtain an estimate of $\beta$, $\hat{\beta}$ using maximum likelihood.
- Our MLE estimator maximizes the **likelihood** conditional on the data

  $$\hat{\beta} = \argmax_\beta \ \log f(\mathbf{y} | \beta, \mathbf{X})$$
  
- This likelihood is defined as the density of the data 
  - But don't confuse it with a density as the input is the **parameter value** and not the **data** itself
  
- Optimization via **numerical techniques**
  - Newton-Raphson
  - Gradient Ascent

:::


# Example: Propensity Scores

- Let's go back to the **Keriakes et. al. (2000)** dataset and play around a bit more with the model structure

```{r, warning=F, message=F, echo=T}
# Read in the dataset
pci <- read_csv("data/pci.csv")
```

- We want to predict treatment: `abcix` using a mix of discrete and continuous covariates...
  - `stent` - Coronary stent deployment; binary indicator
  - `height` - Height in centimeters; numeric integer
  - `female` - Female gender; binary indicator
  - `diabetic` - Diabetes mellitus diagnosis; binary indicator
  - `acutemi` - Acute myocardial infarction within the previous 7 days; binary indicator
  - `ejecfrac` - Left ejection fraction; numeric integer
  - `ves1proc` - Number of vessels involved in the patient’s initial PCI procedure; numeric integer
  

# Example: Propensity Scores

- First, let's just illustrate why OLS is going to not work well for the task of obtaining **fitted values** when we don't have a saturated model.

```{r, warning=F, message=F, echo=T}
ols_reg <- lm(abcix ~ stent*(female + height + diabetic + acutemi + ejecfrac + ves1proc), data=pci)
pci$ols_fitted <- predict(ols_reg)
```

# Example: Propensity Scores

```{r, message=F}
pci %>% ggplot(aes(x=ols_fitted)) + geom_histogram() + xlab("Estimated Probability of abciximab use\nLinear regression") + theme_minimal()
```
  

# Example: Propensity Scores

- Let's fit the logistic regression

```{r, warning=F, message=F, echo=T}
pscore_reg <- glm(abcix ~ stent*(female + height + diabetic + acutemi + ejecfrac + ves1proc), data=pci, 
                  family=binomial(link="logit"))
pci$prob_treat <- predict(pscore_reg, type="response")
```

# Example: Propensity Scores

```{r, message=F}
pci %>% ggplot(aes(x=prob_treat)) + geom_histogram() + xlab("Estimated Probability of abciximab use\nLogit regression") + theme_minimal()
```

# Validating binary classifiers

::: {.incremental}

- We can compute the classic measures of "error" that we're used to (e.g. in-sample or out-of-sample mean squared error).
- But binary classifiers are often visualized using **Receiver Operating Characteristic** (ROC) curves.
- The logistic regression outputs a **continuous** estimate of $Pr(Y = 1 | X)$
  - But converting that to an actual 0/1 **prediction** requires setting some threshold probability.
  
- If we set the threshold low, that guarantees we'll get every **true positive**
  - But we'll get a lot of **false postiives** too
  
- The **ROC** curve captures this trade-off. Different points reflect different classification thresholds
  - **x-axis** - false positive rate
  - **y-axis** - true positive rate

:::

# ROC Curve

- A recent implementation of this exists in the `plotROC` R package

```{r, echo=T}
library(plotROC)
in_sample_roc <- pci %>% ggplot(aes(d = abcix, m = prob_treat)) + geom_roc() + theme_minimal() + 
  labs(x = "False Positive Rate", y = "True Positive Rate") + geom_abline(lty=2, col="indianred")
```

# ROC Curve

```{r}
in_sample_roc 
```

# ROC Curve

- What does a **purely random** classifier look like?

```{r, echo=T}
set.seed(537)
pci$random_class <- runif(nrow(pci))
random_roc <- pci %>% ggplot(aes(d = abcix, m = random_class)) + geom_roc() + theme_minimal() + 
  labs(x = "False Positive Rate", y = "True Positive Rate") + geom_abline(lty=2, col="indianred")
```

# ROC Curve

- It's the 45 degree line!

```{r}
random_roc
```


# ROC Curve

- What does a **perfect** classifier look like?

```{r, echo=T}
true_roc <- pci %>% ggplot(aes(d = abcix, m = abcix)) + geom_roc() + theme_minimal() + 
  labs(x = "False Positive Rate", y = "True Positive Rate") + geom_abline(lty=2, col="indianred")
```

# ROC Curve

- It's the "box"

```{r}
true_roc
```

# ROC Curve

- A predictive model's ROC curve will typically be between those two extremes
  - A common single-variable summary of predictive quality is therefore the **area under curve** (AUC)
  
- What's the AUC of our model in-sample?

```{r, echo=T}
calc_auc(in_sample_roc)
```

# Cross-validation

- We usually don't want to focus too much on **in-sample** fit
  - A more flexible model will always improve in-sample fit (overfitting)
  
- Let's implement a 5-fold cross-validation approach

```{r, echo=T}
set.seed(537)
pci_shuffled <- pci[sample(1:nrow(pci)),]
pci_shuffled$fold <- cut(seq(1:nrow(pci)), breaks = 5, labels = FALSE)
pci_shuffled$prob_treat_cv <- NA
for (f in 1:5){
  pscore_reg_fold <- glm(abcix ~ stent*(female + height + diabetic + acutemi + ejecfrac + ves1proc), data=pci_shuffled %>% filter(fold != f), 
                  family=binomial(link="logit"))
  pci_shuffled$prob_treat_cv[pci_shuffled$fold == f] <- predict(pscore_reg_fold, newdata = pci_shuffled %>% filter(fold == f), type="response")
}
```

# Cross-validation

```{r, echo=T}
oos_roc <- pci_shuffled %>% ggplot(aes(d = abcix, m = prob_treat)) + geom_roc() + geom_roc(aes(m = prob_treat_cv), col="dodgerblue") + theme_minimal() + 
  labs(x = "False Positive Rate", y = "True Positive Rate") + geom_abline(lty=2, col="indianred")
```

# Cross-validation

```{r, echo=T}
oos_roc
```

