---
title: "Week 14: Big Regressions"
subtitle: "PS 818 - Statistical Models"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "December 1, 2025"
date-format: "MMMM D, YYYY"
editor: 
  render-on-save: true
format: 
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 818 - University of Wisconsin - Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute: 
  cache: false
code:
  execute: 
    - r
    - python
embed-resources: true
---


$$
  \require{cancel}
  \DeclareMathOperator*{\argmin}{arg\,min}
$$


$$
 \DeclareMathOperator*{\argmax}{arg\,max}
$$


```{r, echo = F, message=F, warnings=F}
library(tidyverse)
library(haven)
library(estimatr)
library(knitr)
library(splines)
library(lubridate)

options(digits=3)
knitr::opts_chunk$set(echo = T)
set.seed(53703)
```


# This Week

$$
  \DeclareMathOperator*{\argmax}{arg\,max}
$$

::: {.incremental}

# This Week

- Analyses with **big data**
  - Data that cannot fit entirely into computer **memory** (RAM)
  
- General principles for handling large data files

- **Stochastic** optimization methods
  - Stochastic gradient descent for maximum likelihood models

:::

# Review: Generalized Linear Models

::: {.incremental}

- Generalized linear models (GLMs) have three components:
  1) A parametric distribution on $Y_i|X_i$ ("stochastic component")
  2) A linear predictor: $\eta_i = X_i^{\prime}\beta = \beta_0 + \beta_1X_{i1} + \beta_2X_{i2} + \dotsc \beta_kX_{ik}$ ("systematic component")
  3) A link function $g()$ applied to the CEF $E[Y_i|X_i]$ that yields the linear predictor

      $$g(E[Y_i | X_i]) = \eta_i$$

  - Alternatively, we can write the CEF in terms of the "inverse-link" $g^{-1}()$ applied to the linear predictor

      $$E[Y_i | X_i] = g^{-1}(\eta_i)$$

:::

# Review: Logistic regression

::: {.incremental}

- The "logit" or "logistic" GLM models the **log-odds** of a binary outcome as a function of the linear predictor $X_i^{\prime}\beta$

  $$Y_i \underset{\text{i.i.d}}{\sim} \text{Bernoulli}(\pi_i)$$

  $$E[Y_i | X_i] = Pr(Y_i = 1 | X_i) = \pi_i$$

  $$\log\bigg(\frac{\pi_i}{1-\pi_i}\bigg) = X_i^{\prime}\beta$$

- Alternatively, this is written in terms of the "inverse-link" function (the logistic function) that relates $\pi_i$ to $g^{-1}(X_i^{\prime}\beta)$

  $$\pi_i = \frac{\exp(X_i^{\prime}\beta)}{1 + \exp(X_i^{\prime}\beta)} = \frac{1}{1 + \exp(-X_i^{\prime}\beta)}$$

:::

# Review: Logistic regression

- The logistic function maps inputs on $\mathbb{R}$ to $(0, 1)$

```{r, echo=F, fig.width=5, fig.height = 3, fig.align="center"}
logistic <- function(x) 1/(1 + exp(-x))
logistic_curve <- ggplot() + xlim(-5, 5) + ylim(0,1) + geom_function(fun=logistic) + theme_bw() + ylab(expression(pi)) + xlab(expression(paste("X",beta))) + geom_vline(xintercept=0, lty=2) + ggtitle("Logistic function")
logistic_curve
```

# Review: Logistic regression

::: {.incremental}

- Recall the log-likelihood for the logistic regression

$$\log f(\mathbf{y} | \beta, \mathbf{X}) = \sum_{i=1}^{n} \left[ y_i (X_i^{\prime} \beta) - \log(1 + \exp(X_i^{\prime}  \beta)) \right]$$

- We can use the chain rule to get the gradient

$$\nabla \log f(\mathbf{y} | \beta, \mathbf{X}) = \sum_{i=1}^{n} \left[ y_i X_i  - \frac{\exp(X_i^\prime\beta)}{1 + \exp(X_i^\prime\beta)}X_i \right]$$

- Using the definition of the "predicted probabilities" $\pi_i$

$$\nabla \log f(\mathbf{y} | \beta, \mathbf{X}) = \sum_{i=1}^{n} (y_i - \pi_i)X_i$$

:::

# Review: Newton-Raphson

::: {.incremental}

- We obtain an estimate of $\beta$, $\hat{\beta}$ using maximum likelihood.
- Our MLE estimator is:

  $$\hat{\beta} = \argmax_\beta \ \log f(\mathbf{y} | \beta, \mathbf{X})$$

- Recall our score function is the gradient of the log-likelihood

  $$S(\beta) = \nabla \log f(\mathbf{y} | \beta, \mathbf{X}) = \begin{bmatrix} \frac{\partial}{\partial \beta_0} \log f(\mathbf{y} | \beta, \mathbf{X}) \\
\frac{\partial}{\partial \beta_1} \log f(\mathbf{y} | \beta, \mathbf{X}) \\
\vdots \\
\frac{\partial}{\partial \beta_k} \log f(\mathbf{y} | \beta, \mathbf{X})  \end{bmatrix}$$

- The likelihood is convex, so finding the maximum equates to solving for the value of $\beta$ that sets $S(\beta) = 0$
  - A common numerical method is **Newton-Raphson**

:::

# Review: Newton-Raphson

::: {.incremental}

- Given some initial starting values $\hat{\beta}^{(0)}$, our update steps $\hat{\beta}^{(t+1)}$ proceed deterministically until convergence using previous values $\hat{\beta}^{(t)}$

  $$\hat{\beta}^{(t+1)} =  \hat{\beta}^{(t)} + \bigg(-\mathbf{H}^{-1}(\hat{\beta}^{(t)})\bigg) \nabla \log f(\mathbf{y} | \hat{\beta}^{(t)}, \mathbf{X})$$
  
- What is this information?
  - The **gradient** of the log-likelihood: $\nabla \log f(\mathbf{y} | \hat{\beta}^{(t)}, \mathbf{X})$
  - The (negative) inverse **hessian** (step size)
- Why is this hard? Why don't we do this for "machine learning"
  - Because inverting large matrices is computationally costly
  
:::

# Gradient Ascent

::: {.incremental}

- Instead of changing the "step size" in a data-dependent manner, what if we just fixed it 
- The **gradient ascent** algorithm would instead use update steps

  $$\hat{\beta}^{(t+1)} =  \hat{\beta}^{(t)} + \gamma_{t}\nabla \log f(\mathbf{y} | \hat{\beta}^{(t)}, \mathbf{X})$$

- We refer to $\gamma = \{\gamma_0, \gamma_1, \gamma_2, \dotsc \} > 0$ as our "learning rate schedule"
  - Too big, we overshoot
  - Too small, we don't explore the likelihood
- Often just pick a constant rate that doesn't depend on the iteration $\gamma_0 = \gamma_1 = \gamma_2 = \dotsc$
  - Alternatively, set some degree of **decay** (e.g. exponential decay $\gamma_{t} = \gamma_0 \times e^{-kt}$)

:::

# Large datasets with `data.table`

- We're going to look at data management with the R package `data.table`

```{r, message=F, warning=F}
library(data.table)
```

- Why use `data.table` instead of the standard `read_csv` or `read.csv`?
  - It handles **very large** datasets much better!
  - Faster file-IO, "chunking" of datasets that can't fit into memory, operations directly on datasets (avoid copying and duplicating memory), etc...
- If you genuinely have an **extremely large** dataset, you'll probably be working with a **relational database**
  - Typical interaction via the **SQL** language. 
  - If you know `tidy` verbs, you can learn SQL easily. Also similar motivation for `data.table` syntax
  - Outside the scope of this class at this moment (but core principles remain).

# Example: NC Voter File

- Let's do a simple predictive model:
  - What is the probability that someone voted in the **2024 General Election** given known information (age, party, gender, race) as well as their prior voting history.
- We'll work with the most current version of the North Carolina Voter File and the Voter History file (since it's free to download)
  - Millions of records! 
  - Unfortunately...two separate databases that need to be merged.
  - One is a **6 GB** text file, the other is **4 GB** text file
  
# Example: NC Voter File

- Let's load the data into R with `fread()`

```{r, echo=T}
ncvhis <- fread("data/ncvhis_Statewide.txt") # Reads in the whole voter history file
ncvhis <- ncvhis[,c("ncid","election_lbl", "voting_method")] # We just want the ID of the voter and which elections they voted in
ncvhis <- ncvhis[!duplicated(ncvhis, by = c("ncid", "election_lbl"))] # Get rid of duplicates
gc() # "Garbage collect" to reallocate memory.
```

# Example: NC Voter File

- We want to make the dataset *wide* so that we can merge the elections where the voter voted with the actual voter file.
  - Each column should be an election - the content should be an indicator for "voted/didn't vote"

```{r, echo=T}
ncvhis_wide <- dcast(ncvhis, ncid ~ election_lbl, value.var = "voting_method") # Equivalent of pivot.wider
rm(ncvhis) # Get rid of the old data.table
gc() # Garbage collect to free up memory!
head(ncvhis_wide) # What does the new dataset look like?
```

# Example: NC Voter File

- Let's grab the last five general elections

```{r, echo=T}
ncvhis_wide <- ncvhis_wide[,c("ncid","11/05/2024", "11/08/2022", "11/03/2020", "11/06/2018", "11/08/2016")]
ncvhis_wide[,`11/05/2024` := fifelse(!is.na(`11/05/2024`), 1, 0)] # Modify the columns *directly*
ncvhis_wide[,`11/08/2022` := fifelse(!is.na(`11/08/2022`), 1, 0)] 
ncvhis_wide[,`11/03/2020` := fifelse(!is.na(`11/03/2020`), 1, 0)] 
ncvhis_wide[,`11/06/2018` := fifelse(!is.na(`11/06/2018`), 1, 0)] 
ncvhis_wide[,`11/08/2016` := fifelse(!is.na(`11/08/2016`), 1, 0)]
setnames(ncvhis_wide, old = c("11/05/2024", "11/08/2022", "11/03/2020", "11/06/2018", "11/08/2016"), 
         new = c("voted2024", "voted2022", "voted2020", "voted2018", "voted2016"))
```

# Example: NC Voter File

- Now we need to merge it to the voter file. Let's load up the voter file!

```{r}
ncvoter <- fread("data/ncvoter_Statewide.txt")
ncvoter <- ncvoter[voter_status_desc != "REMOVED"&voter_status_desc != "DENIED"]
```

- We'll need to now merge based on `ncid`

# Example: NC Voter File

```{r}
ncvoter <- merge(ncvoter, ncvhis_wide, by="ncid", all.x=T, all.y=F)
rm(ncvhis_wide) # Get rid of the old file!
ncvoter <- ncvoter[,c('ncid','voted2024', 'voted2020','voted2016','voted2022','voted2018','birth_year','gender_code','ethnic_code', 'party_cd', 'race_code')]
gc()
```

- Need to make any voter entries that didn't match into **non-voters**

```{r}
ncvoter[is.na(voted2024), voted2024 := 0] # Again, modifying the columns *directly*
ncvoter[is.na(voted2022), voted2022 := 0]
ncvoter[is.na(voted2020), voted2020 := 0]
ncvoter[is.na(voted2018), voted2018 := 0]
ncvoter[is.na(voted2016), voted2016 := 0]
```

- How many voted?

```{r}
mean(ncvoter$voted2024)
```

# Trying to fit logistic regression

- Can we actually fit `glm()` to a dataset that takes up more than half of my laptop's memory?
  - We can - this implementation does work - but **only just**

```{r}
voting_reg <- glm(voted2024 ~ voted2020 + voted2016 + voted2022 + voted2018 + 
                    I(birth_year <= 1964) + I(birth_year > 1964&birth_year <= 1980) + 
                    I(birth_year > 1997) + I(gender_code == "M") + I(ethnic_code == "HL") + 
                    I(race_code == "W") + I(party_cd == "REP") + 
                    I(party_cd == "UNA") + I(party_cd == "LIB") + I(party_cd == "GRE"), data=ncvoter, family=binomial(link="logit"))
gc()
```

# Trying to fit logistic regression

```{r}
summary(voting_reg)
```

# Sparse Matrices

- Let's consider how large the data matrix is when I use `model.matrix` for the regression.

```{r}
big.design.matrix <- model.matrix(voted2024 ~ voted2020 + voted2016 + voted2022 + voted2018 + I(birth_year <= 1964) + I(birth_year > 1964&birth_year <= 1980) + I(birth_year > 1997) + I(gender_code == "M") + I(ethnic_code == "HL") + I(race_code == "W") + I(party_cd == "REP") + I(party_cd == "UNA") + I(party_cd == "LIB") + I(party_cd == "GRE"), data=ncvoter)
print(object.size(big.design.matrix), units="MB")
rm(big.design.matrix)
gc()
```

# Sparse Matrices

- A more memory-efficient way of storing things is in a **sparse** matrix.
  - Especially when most entries are **zero**. We can just store a list of all of the indices that are equal to $1$
  - Nice implementation in `MatrixModels`
  - Lots of libraries are designed around using **sparse** matrices - consider these as a solution if you are running analyses that involve matrices with lots of zeroes.
  
```{r}
library(MatrixModels)
sparse.design.matrix <- model.Matrix(voted2024 ~ voted2020 + voted2016 + voted2022 + voted2018 + I(birth_year <= 1964) + I(birth_year > 1964&birth_year <= 1980) + I(birth_year > 1997) + I(gender_code == "M") + I(ethnic_code == "HL") + I(race_code == "W") + I(party_cd == "REP") + I(party_cd == "UNA") + I(party_cd == "LIB") + I(party_cd == "GRE"), data=ncvoter, sparse=T)
print(object.size(sparse.design.matrix), units="MB")
rm(sparse.design.matrix)
gc()
```

# Stochastic Gradient Descent

::: {.incremental}

- Let's return to our **algorithm** for constructing estimates of $\beta$
  - In conventional ('batch') gradient ascent, we used the **entire** dataset to compute the gradient.
  - But what if we can't load the entire dataset into memory!
  
- **Solution**: "Online" learning
  - What if we computed the gradient updates using a randomly selected **single observation**?
  
$$\hat{\beta}^{(t+1)} =  \hat{\beta}^{(t)} + \gamma_{t+1} \nabla \log f(y_{t+1} | \hat{\beta}^{(t)}, X_{t+1})$$

- **Robbins and Munro (1951)** showed that this will converge as the number of iterations goes to infinity subject to conditions on the learning rate schedule
  - $\sum_{t=1}^{\infty} \gamma_{t} = \infty$, $\sum_{t=1}^{\infty} \gamma_{t}^2 < \infty$
  
- Can improve performance further if we use "mini-batches" to estimate the gradient (more than one randomly sampled observation).

:::

# Stochastic Gradient Descent

- A very nice implementation is in the R `sgd` package.
  - Handles all the usual formula syntax stuff for a large number of statistical models.
  
```{r}
library(sgd)
```

- Let's implement standard `sgd` for the logistic GLM from before and see how close it gets

```{r}
set.seed(53703)
voting_reg_sgd <- sgd(voted2024 ~ voted2020 + voted2016 + voted2022 + voted2018 + I(birth_year <= 1964) + I(birth_year > 1964&birth_year <= 1980) + I(birth_year > 1997) + I(gender_code == "M") + I(ethnic_code == "HL") + I(race_code == "W") + I(party_cd == "REP") + I(party_cd == "UNA") + I(party_cd == "LIB") + I(party_cd == "GRE"), data=ncvoter, model = "glm", model.control = list(family=binomial(link="logit")),
                      sgd.control = list(method="sgd"))
voting_reg_sgd$converged
gc()
```

# Stochastic Gradient Descent

```{r}
coef(voting_reg_sgd)
```

```{r}
coef(voting_reg)
```

# Stochastic Gradient Descent

- Let's check the in-sample predictions (fitted data)

```{r}
fitted(voting_reg_sgd)[1:10]
```

```{r}
fitted(voting_reg)[1:10]
```

- Remember, the guarantees of SGD are **asymptotic**
  - Actual performance in a finite number of iterations is highly sensitive to the specified **learning rate**
  
# Stochastic Gradient Descent

- Lots of recent work on improving these methods (e.g. **adaptive** gradient methods, **momentum**)
  - Requires some additional parameters to keep in memory (previous values of the gradient)
  
- Outside the scope of this course, but important to be aware.
  - Currently the most popular approach for initial training of neural networks is the **Adam** optimizer.
  
# Stochastic Gradient Descent

- Let's see how well we at least do with the "adaptive" learning rate?

```{r}
set.seed(53703)
voting_reg_ada_sgd <- sgd(voted2024 ~ voted2020 + voted2016 + voted2022 + voted2018 + I(birth_year <= 1964) + I(birth_year > 1964&birth_year <= 1980) + I(birth_year > 1997) + I(gender_code == "M") + I(ethnic_code == "HL") + I(race_code == "W") + I(party_cd == "REP") + I(party_cd == "UNA") + I(party_cd == "LIB") + I(party_cd == "GRE"), data=ncvoter, model = "glm", model.control = list(family=binomial(link="logit")),
                      sgd.control = list(method="sgd", lr="adagrad"))
voting_reg_ada_sgd$converged
gc()
```

# Stochastic Gradient Descent

```{r}
coef(voting_reg_ada_sgd)
```

```{r}
coef(voting_reg)
```

# Stochastic Gradient Descent

- Seems to be doing **worse** - we missed the intercept!

```{r}
fitted(voting_reg_ada_sgd)[1:10]
```

```{r}
fitted(voting_reg)[1:10]
```

# Stochastic Gradient Descent

- **Toulis, Tran and Airoldi (2015)** introduce a clever technique that tries to reduce sensitivity to learning rate parameter specification by defining the update steps **implicitly** and using methods for solving implicit equations.
  
  $$\hat{\beta}^{(t+1)} =  \hat{\beta}^{(t)} + \gamma_{t+1} \nabla \log f(y_{t+1} | \hat{\beta}^{(t+1)}, X_{t+1})$$


# Stochastic Gradient Descent

- How well does the **averaged-implicit** SGD work here?

```{r}
set.seed(53703)
voting_reg_isgd <- sgd(voted2024 ~ voted2020 + voted2016 + voted2022 + voted2018 + I(birth_year <= 1964) + I(birth_year > 1964&birth_year <= 1980) + I(birth_year > 1997) + I(gender_code == "M") + I(ethnic_code == "HL") + I(race_code == "W") + I(party_cd == "REP") + I(party_cd == "UNA") + I(party_cd == "LIB") + I(party_cd == "GRE"), data=ncvoter, model = "glm", model.control = list(family=binomial(link="logit")),
                      sgd.control = list(method="ai-sgd", lr="adagrad"))
voting_reg_isgd$converged
gc()
```

# Stochastic Gradient Descent

```{r}
coef(voting_reg_isgd)
```

```{r}
coef(voting_reg)
```

# Stochastic Gradient Descent

- Overall not bad - and that's only after a few passes over the dataset!

```{r}
fitted(voting_reg_isgd)[1:10]
```

```{r}
fitted(voting_reg)[1:10]
```

